---
title: "BLASTp Results Filtering and Best-Hit Curation"
author: "David Alberto García Estrada"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This document filters BLASTp results from the custom protein database search and
selects the **single best hit per Query–Organism pair**. The curation strategy
ensures one representative sequence per organism per gene family enters
downstream phylogenetic analysis, preventing organisms with paralog expansions
from dominating the dataset.

Input: 15-column tabular BLASTp output (`outfmt 6`) with columns:
`qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore qcovs qlen slen`

Best-hit selection priority (within each Query–Organism group):

1. **Bit score** (highest) — alignment quality
2. **qcovs** (highest) — fraction of the query aligned
3. **pident** (highest) — sequence identity
4. **E-value** (lowest) — statistical significance

---

# 1. Configuration

```{r configuration}
library(tidyverse)

# =============================================================================
# USER CONFIGURATION — edit these paths before running
# =============================================================================

# Path to the filtered BLASTp results (output of blastp_batch.sh)
INPUT_BLAST_FILE <- "results/blast/filtered/<protein_name>_blastp_filtered.tsv"  # <-- CHANGE THIS

# Directory where output files will be written
OUTPUT_DIRECTORY <- "results/blast/unique/<protein_name>"                        # <-- CHANGE THIS

# Organism extraction method — must match the header format of hormoneDB.fasta.
# Options:
#   "taxid"        — TaxID prefix before the first "|"  (e.g., "105231")
#                    Use this when the database was built with prepare_blastdb.sh
#   "species_code" — UniProt species code (e.g., "KLENI" from A0A1Y1HVA8_KLENI)
#   "os_name"      — Full species name after "OS=" in the header
ORGANISM_METHOD <- "taxid"

# =============================================================================

cat(sprintf("✓ Configuration loaded\n"))
cat(sprintf("  Input file      : %s\n", INPUT_BLAST_FILE))
cat(sprintf("  Output directory: %s\n", OUTPUT_DIRECTORY))
cat(sprintf("  Organism method : %s\n", ORGANISM_METHOD))
```

---

# 2. Data Loading

```{r load_data}
cat(sprintf("\nLoading BLASTp results from '%s'...\n", INPUT_BLAST_FILE))

blast_results <- read_tsv(
  INPUT_BLAST_FILE,
  col_names = TRUE,
  show_col_types = FALSE
)

cat(sprintf("✓ Data loaded: %d hits, %d columns\n",
            nrow(blast_results), ncol(blast_results)))

# Verify expected 15-column extended format
expected_cols <- c("qseqid", "sseqid", "pident", "length", "mismatch",
                   "gapopen", "qstart", "qend", "sstart", "send",
                   "evalue", "bitscore", "qcovs", "qlen", "slen")

missing_cols <- setdiff(expected_cols, colnames(blast_results))
if (length(missing_cols) > 0) {
  cat("\n⚠ WARNING: Missing expected columns:",
      paste(missing_cols, collapse = ", "), "\n")
} else {
  cat("✓ All 15 expected columns present\n")
}

cat("\nPreview (key columns):\n")
blast_results %>%
  select(qseqid, sseqid, pident, evalue, bitscore, qcovs) %>%
  head(3) %>%
  print()
```

---

# 3. Organism Extraction

```{r extract_organism}
cat("\nExtracting organism identifier from sseqid...\n")

# Extract organism ID from the subject sequence header.
# The method must match the header format used when building hormoneDB.fasta:
#   "taxid"        → "105231|tr|A0A1Y1HVA8|..." → "105231"
#   "species_code" → "...|A0A1Y1HVA8_KLENI ..." → "KLENI"
#   "os_name"      → "... OS=Klebsormidium nitens OX=..." → "Klebsormidium nitens"
blast_results <- blast_results %>%
  mutate(organism = case_when(
    ORGANISM_METHOD == "taxid" ~
      str_extract(sseqid, "^[^|]+"),
    ORGANISM_METHOD == "species_code" ~
      str_extract(sseqid, "_([A-Z0-9]+)\\s", group = 1),
    ORGANISM_METHOD == "os_name" ~
      str_extract(sseqid, "OS=([^=]+?)(?:OX=|GN=|PE=|$)", group = 1) %>%
      str_trim(),
    TRUE ~ str_extract(sseqid, "^[^|]+")
  ))

cat("\nExtraction examples (first 5 unique organisms):\n")
blast_results %>%
  select(sseqid, organism) %>%
  distinct(organism, .keep_all = TRUE) %>%
  head(5) %>%
  print(width = Inf)

num_organisms <- n_distinct(blast_results$organism)
num_queries   <- n_distinct(blast_results$qseqid)
cat(sprintf("\n✓ Unique organisms: %d\n", num_organisms))
cat(sprintf("✓ Unique queries  : %d\n", num_queries))

# Flag headers that did not yield a valid organism ID
num_na <- sum(is.na(blast_results$organism))
if (num_na > 0) {
  cat(sprintf("\n⚠ WARNING: %d hits with no extractable organism ID\n", num_na))
  cat("  Problematic sseqid examples:\n")
  blast_results %>%
    filter(is.na(organism)) %>%
    select(sseqid) %>%
    head(3) %>%
    print()
  cat("  → Check ORGANISM_METHOD or the sseqid format in your database\n")
}
```

---

# 4. Pre-filtering Exploratory Analysis

```{r exploratory_analysis}
cat("\nQuery–Organism hit distribution (before filtering)...\n")

# Count hits per Query–Organism combination to understand redundancy
duplicates_summary <- blast_results %>%
  group_by(qseqid, organism) %>%
  summarise(num_hits = n(), .groups = "drop") %>%
  arrange(desc(num_hits))

cat(sprintf("  Total Query–Organism combinations : %d\n", nrow(duplicates_summary)))
cat(sprintf("  Combinations with 1 hit           : %d\n",
            sum(duplicates_summary$num_hits == 1)))
cat(sprintf("  Combinations with multiple hits   : %d\n",
            sum(duplicates_summary$num_hits > 1)))

cat("\nTop 10 combinations with most hits (will be reduced to 1):\n")
duplicates_summary %>%
  filter(num_hits > 1) %>%
  head(10) %>%
  print()

# Show a concrete example of multiple hits being collapsed
if (sum(duplicates_summary$num_hits > 1) > 0) {
  example_case <- duplicates_summary %>%
    filter(num_hits > 1) %>%
    slice(1)

  cat(sprintf("\nExample — Query '%s' + Organism '%s' (%d hits):\n",
              example_case$qseqid,
              example_case$organism,
              example_case$num_hits))

  blast_results %>%
    filter(qseqid == example_case$qseqid,
           organism == example_case$organism) %>%
    select(sseqid, pident, evalue, bitscore, qcovs, length) %>%
    arrange(desc(bitscore)) %>%
    print()

  cat("→ Only the first row (highest bitscore) will be retained\n")
}
```

---

# 5. Best-Hit Filtering

```{r process_data}
cat("\n", rep("=", 70), "\n", sep = "")
cat("FILTERING: ONE BEST HIT PER QUERY–ORGANISM\n")
cat(rep("=", 70), "\n", sep = "")

# Rationale: grouping by organism (not by sequence ID) ensures that organisms
# with paralog expansions contribute only one representative to the alignment,
# maintaining balanced taxon sampling in the phylogenetic analysis.

best_hits_per_query_organism <- blast_results %>%

  # Group by query AND organism — each pair is treated as an independent unit
  group_by(qseqid, organism) %>%

  # Sort within each group by decreasing quality metrics:
  #   1. bitscore (desc): alignment quality
  #   2. qcovs    (desc): query coverage
  #   3. pident   (desc): percent identity
  #   4. evalue   (asc) : statistical significance
  arrange(desc(bitscore), desc(qcovs), desc(pident), evalue,
          .by_group = TRUE) %>%

  # Retain only the top-ranked hit per group
  slice(1) %>%
  ungroup()

# Reduction statistics
cat(sprintf("\n✓ Filtering complete\n"))
cat(sprintf("  Original hits       : %7d\n", nrow(blast_results)))
cat(sprintf("  After filtering     : %7d\n", nrow(best_hits_per_query_organism)))
cat(sprintf("  Redundant hits removed: %7d (%.1f%%)\n",
            nrow(blast_results) - nrow(best_hits_per_query_organism),
            100 * (1 - nrow(best_hits_per_query_organism) / nrow(blast_results))))

cat(sprintf("\n  Queries processed  : %d\n",
            n_distinct(best_hits_per_query_organism$qseqid)))
cat(sprintf("  Unique organisms   : %d\n",
            n_distinct(best_hits_per_query_organism$organism)))

# Per-query organism distribution
cat("\nOrganism count per query:\n")
organism_distribution <- best_hits_per_query_organism %>%
  count(qseqid, name = "num_organisms") %>%
  arrange(desc(num_organisms))
print(organism_distribution)

cat(sprintf("\n  Mean   : %.1f organisms/query\n",
            mean(organism_distribution$num_organisms)))
cat(sprintf("  Median : %.0f\n",
            median(organism_distribution$num_organisms)))
cat(sprintf("  Range  : %d – %d\n",
            min(organism_distribution$num_organisms),
            max(organism_distribution$num_organisms)))
```

---

# 6. Quality Verification

```{r quality_check}
cat("\n", rep("=", 70), "\n", sep = "")
cat("QUALITY CHECKS\n")
cat(rep("=", 70), "\n", sep = "")

# Test 1: Uniqueness — no duplicated Query–Organism pairs
cat("\nTest 1: Query–Organism uniqueness...\n")
duplicates_check <- best_hits_per_query_organism %>%
  group_by(qseqid, organism) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(count > 1)

if (nrow(duplicates_check) > 0) {
  cat("❌ FAIL: Duplicated Query–Organism pairs found:\n")
  print(duplicates_check)
  cat("  → This should not occur. Review the filtering code.\n")
} else {
  cat("✅ PASS: Each Query–Organism pair has exactly 1 hit\n")
}

# Test 2: Value ranges — pident and qcovs must be in [0, 100]
cat("\nTest 2: Value ranges...\n")

invalid_pident <- sum(best_hits_per_query_organism$pident < 0 |
                      best_hits_per_query_organism$pident > 100)
if (invalid_pident > 0) {
  cat(sprintf("  ⚠ %d hits with pident outside [0, 100]\n", invalid_pident))
} else {
  cat("  ✅ pident: all values in valid range [0, 100]\n")
}

invalid_qcovs <- sum(best_hits_per_query_organism$qcovs < 0 |
                     best_hits_per_query_organism$qcovs > 100)
if (invalid_qcovs > 0) {
  cat(sprintf("  ⚠ %d hits with qcovs outside [0, 100]\n", invalid_qcovs))
} else {
  cat("  ✅ qcovs: all values in valid range [0, 100]\n")
}

# Test 3: Detailed example for one query
cat("\nTest 3: Detailed example for the first query...\n")
example_query <- best_hits_per_query_organism$qseqid[1]
cat(sprintf("  Query: '%s'\n", example_query))
cat(sprintf("  Organisms found: %d\n\n",
            sum(best_hits_per_query_organism$qseqid == example_query)))

best_hits_per_query_organism %>%
  filter(qseqid == example_query) %>%
  select(organism, sseqid, pident, qcovs, evalue, bitscore) %>%
  arrange(desc(bitscore)) %>%
  head(10) %>%
  print(width = Inf)

# Test 4: Overall quality statistics
cat("\nTest 4: Quality statistics of selected hits...\n")
quality_stats <- best_hits_per_query_organism %>%
  summarise(
    avg_pident    = mean(pident),
    median_pident = median(pident),
    min_pident    = min(pident),
    avg_qcovs     = mean(qcovs),
    median_qcovs  = median(qcovs),
    min_qcovs     = min(qcovs),
    avg_bitscore  = mean(bitscore),
    median_evalue = median(evalue)
  )

cat(sprintf("  Identity  — mean: %.2f%%  median: %.2f%%  min: %.2f%%\n",
            quality_stats$avg_pident,
            quality_stats$median_pident,
            quality_stats$min_pident))
cat(sprintf("  Coverage  — mean: %.2f%%  median: %.2f%%  min: %.2f%%\n",
            quality_stats$avg_qcovs,
            quality_stats$median_qcovs,
            quality_stats$min_qcovs))
cat(sprintf("  Bitscore  — mean: %.2f\n",  quality_stats$avg_bitscore))
cat(sprintf("  E-value   — median: %.2e\n", quality_stats$median_evalue))
```

---

# 7. Output Files

```{r save_results}
cat("\n", rep("=", 70), "\n", sep = "")
cat("SAVING RESULTS\n")
cat(rep("=", 70), "\n", sep = "")

if (!dir.exists(OUTPUT_DIRECTORY)) {
  dir.create(OUTPUT_DIRECTORY, recursive = TRUE)
  cat(sprintf("✓ Output directory created: %s\n", OUTPUT_DIRECTORY))
}

# File 1: Consolidated table — all queries, one row per Query–Organism
consolidated_path <- file.path(OUTPUT_DIRECTORY, "ALL_queries_best_hits.tsv")
write_tsv(best_hits_per_query_organism, consolidated_path)
cat(sprintf("\n✓ [1/3] Consolidated table:\n"))
cat(sprintf("        %s\n", consolidated_path))
cat(sprintf("        %d rows (1 hit per Query–Organism)\n",
            nrow(best_hits_per_query_organism)))

# File 2: Individual tables per query
cat(sprintf("\n✓ [2/3] Individual tables per query:\n"))
num_files_saved <- 0
best_hits_per_query_organism %>%
  group_by(qseqid) %>%
  group_walk(~ {
    query_name    <- .y$qseqid
    safe_filename <- str_replace_all(query_name, "[^[:alnum:]_-]", "_")
    output_path   <- file.path(OUTPUT_DIRECTORY, paste0(safe_filename, ".tsv"))
    write_tsv(.x, output_path)
    num_files_saved <<- num_files_saved + 1
    cat(sprintf("        %s (%d organisms)\n",
                basename(output_path), nrow(.x)))
  })

# File 3: Per-query summary statistics
summary_stats <- best_hits_per_query_organism %>%
  group_by(qseqid) %>%
  summarise(
    num_organisms    = n(),
    avg_identity     = round(mean(pident),    2),
    median_identity  = round(median(pident),  2),
    avg_coverage     = round(mean(qcovs),     2),
    median_coverage  = round(median(qcovs),   2),
    avg_evalue       = mean(evalue),
    avg_bitscore     = round(mean(bitscore),  2),
    best_bitscore    = round(max(bitscore),   2),
    .groups = "drop"
  ) %>%
  arrange(desc(num_organisms))

summary_path <- file.path(OUTPUT_DIRECTORY, "summary_statistics.tsv")
write_tsv(summary_stats, summary_path)
cat(sprintf("\n✓ [3/3] Summary statistics:\n"))
cat(sprintf("        %s\n", summary_path))

cat("\n", rep("=", 70), "\n", sep = "")
```

---

# 8. Summary and Visualizations

```{r final_summary, fig.width=12, fig.height=8}
cat("\n")
cat("╔", rep("═", 68), "╗\n", sep = "")
cat("║", sprintf("%68s", "ANALYSIS SUMMARY"), "║\n", sep = "")
cat("╚", rep("═", 68), "╝\n\n", sep = "")

cat(sprintf("  Queries analyzed          : %d\n",
            n_distinct(best_hits_per_query_organism$qseqid)))
cat(sprintf("  Unique organisms found    : %d\n",
            n_distinct(best_hits_per_query_organism$organism)))
cat(sprintf("  Final hits (1 per Q–O)    : %d\n",
            nrow(best_hits_per_query_organism)))
cat(sprintf("  Original hits             : %d\n",
            nrow(blast_results)))
cat(sprintf("  Redundancy removed        : %.1f%%\n",
            100 * (1 - nrow(best_hits_per_query_organism) / nrow(blast_results))))
cat(sprintf("\n  Output directory          : %s\n", OUTPUT_DIRECTORY))
cat(sprintf("  Consolidated table        : ALL_queries_best_hits.tsv\n"))
cat(sprintf("  Individual tables         : %d files\n", num_files_saved))
cat(sprintf("  Summary statistics        : summary_statistics.tsv\n"))
cat("\n✓ Analysis complete\n\n")

library(gridExtra)

# Plot 1: Organism count per query
p1 <- organism_distribution %>%
  ggplot(aes(x = reorder(qseqid, num_organisms), y = num_organisms)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = num_organisms), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Organisms recovered per query",
       subtitle = "Each bar represents one query sequence",
       x = "Query ID",
       y = "Number of organisms") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

# Plot 2: Percent identity distribution
p2 <- best_hits_per_query_organism %>%
  ggplot(aes(x = pident)) +
  geom_histogram(bins = 30, fill = "coral", alpha = 0.7, color = "darkred") +
  geom_vline(aes(xintercept = mean(pident)),
             color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text",
           x = mean(best_hits_per_query_organism$pident) + 5,
           y = Inf, vjust = 2,
           label = sprintf("Mean: %.1f%%",
                           mean(best_hits_per_query_organism$pident)),
           color = "red") +
  labs(title = "Percent identity distribution",
       subtitle = "Selected best hits",
       x = "% Identity",
       y = "Count") +
  theme_minimal()

# Plot 3: Query coverage distribution
p3 <- best_hits_per_query_organism %>%
  ggplot(aes(x = qcovs)) +
  geom_histogram(bins = 30, fill = "seagreen", alpha = 0.7, color = "darkgreen") +
  geom_vline(aes(xintercept = mean(qcovs)),
             color = "darkgreen", linetype = "dashed", linewidth = 1) +
  annotate("text",
           x = mean(best_hits_per_query_organism$qcovs) + 5,
           y = Inf, vjust = 2,
           label = sprintf("Mean: %.1f%%",
                           mean(best_hits_per_query_organism$qcovs)),
           color = "darkgreen") +
  labs(title = "Query coverage distribution (qcovs)",
       subtitle = "Fraction of the query sequence aligned with the subject",
       x = "% Query coverage",
       y = "Count") +
  theme_minimal()

# Plot 4: Identity vs coverage scatter
p4 <- best_hits_per_query_organism %>%
  ggplot(aes(x = pident, y = qcovs)) +
  geom_point(alpha = 0.4, color = "purple") +
  geom_density_2d(color = "darkviolet") +
  labs(title = "Identity vs. query coverage",
       subtitle = "High-quality hits cluster in the upper-right corner",
       x = "% Identity",
       y = "% Query coverage") +
  theme_minimal()

print(p1)
print(p2)
print(p3)
print(p4)
```

---

# 9. Per-Query Summary Table

```{r summary_table}
cat("\nPer-query summary:\n\n")
summary_stats %>%
  mutate(avg_evalue = sprintf("%.2e", avg_evalue)) %>%
  print(n = Inf)
```

---

# 10. Technical Notes

**Grouping strategy — why by organism, not by sequence ID:**

```r
# Incorrect: groups by individual sequence — allows multiple hits from the same organism
group_by(qseqid, sseqid)

# Correct: groups by organism — one representative per organism per query
mutate(organism = str_extract(sseqid, "^[^|]+"))
group_by(qseqid, organism)
```

**Organism extraction methods:**

| Method | Pattern | Use case |
|---|---|---|
| `"taxid"` | First field before `\|` | Default; matches `prepare_blastdb.sh` output |
| `"species_code"` | UniProt 5-letter code | Alternative UniProt-style headers |
| `"os_name"` | Text after `OS=` | Full species name from UniProt headers |

**Interpretation:** if a query returns 45 organisms, it means 45 *distinct* organisms, each represented by exactly one sequence — the highest-scoring hit for that organism.

---

# Session Info

```{r session_info}
sessionInfo()
```
